Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-03-06 src/tex/ms.tex
[showyourwork] overleaf sync

\newcommand{\todo}[1]{\textcolor{red}{[#1]}}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory, a coronagraphic mission which would detect rocky planets in the habitable zone, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. The glint signature at scattering phases of $\sim$50--70$^\circ$ would be accessible to the Habitable Worlds Observatory in \textbf{XX} systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in \textbf{YY} systems. The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}

@@

\textcolor{red}{introduce motivation, specs for habitable worlds observatory and the document we got the target list from.}

\textcolor{blue}{Habitable worlds bit}

@@

\textbf{Why glint [Kim]}
% probe of surface conditions; habitability

%\cite{2006astro.ph.10518M} first noted the potential for photometric and polarimetric scattered light phase to detect oceans on terrestrial exoplanets.   --> No it was talked about previous to this, per the paper cited here in fact... "Stam & Hovenier(2006) independently examined the observability of the polarized signatures of an Earth-likeextrasolar planet, including Rayleigh scattering and sea-surface glint. Schneider (2006) had similar ideas." -KB

%\cite{2008Icar..195..927W} simulated photometric and polarimetric phase curves of directly imaged terrestrial planets. They noted that surface oceans would be detectable because they are a) less reflective near full phase, 
% not following what you mean here -KB
%b) more reflective at crescent phases, and c) exhibit a peak in polarization at crescent phases. \cite{Zugger_2010} simulated photometric and polarized phase variations for planets including the effects of partial cloud coverage and ocean waves. Although the presence of clouds and atmospheric Rayleigh scattering weaken the signal of an underlying ocean, they again find that oceans are detectable and distinguishable from other sources of polarization. \cite{2010ApJ...721L..67R} compared observations of the Earth's scattering phase curve to simulations; they found .
%\citet{Zugger_2011} note that ocean glint might be observed in near infrared opacity windows, because of the reduced Rayleigh scattering at long wavelengths. This has the added advantage of minimizing the latitude--albedo effect, which could otherwise be a false-positive for photometric brightening at crescent phases due to glint \citep{2012ApJ...752L...3C}. \cite{2019A&A...626A.129T} simulated the photometric and polarimetric signatures of terrestrial planets with inhomogeneous clouds. \cite{2022A&A...664A.172T}

@@

@@

\section{Methods/simulations}
\subsection{\hwo\ and Stellar Sample}
%\textcolor{blue}{Merge the 2 sections here into "Assumed technical specifications and sample"? I don't think this needs to be 2 subsections}
\hwo\ is still in the early stages of development, and the exact telescope and instrument design is undecided. 
%
In this work, we assume a \SI{6}{\meter} primary mirror, which is the currently favoured design for the observatory.
%
In addition, \hwo\ will utilise a coronagraph to suppress the stellar light and facilitate high contrast observations of their faint exoplanet companions. 
%
The \IWA\ of the coronagraph is currently undecided and will be driven by both the science requirements and technological limitations. 
%In this work, we investigate the scattering phase coverage obtainable for several choices for the \IWA\ of the observatory. 
%We also consider the Outer Working Angle (OWA) but for most targets this is not a limiting factor. 
For ease of comparison, \cref{tab:IWA_OWA} indicates the angular separations corresponding to multiples of $\lambda / D$ for a \SI{6}{\meter} telescope at a wavelength of $\lambda = \SI{-1}{\nano\meter}$ \todo{Update value!}.

@@

%\subsection{Stellar sample and assumed planet parameters} 
\hwo is currently envisioned to observe the stars stated in the NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory.%
\footnote{\url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list comprises $\sim$160 stars, the majority of which are Sun-like dwarfs; 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude of $R_c = 31$ and a contrast floor of \num{2.5e-11}. 
%with an occurrence rate of rocky planets in the optimistic habitable zone to be $\eta_{\oplus}$ = 0.24 \citep{decadal}. 
The corresponding adopted habitable zone limits are a semi-major axis of \SIrange{0.95}{1.67}{\au} for a solar twin, planet sizes between \SIrange{0.8}{1.4}{} Earth radii, and for non-solar stars, scale as square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}.

@@

The coronagraph, while necessary for observing faint exoplanets, can obscure part of the planet's orbit. 
%
As demonstrated in \cref{fig:annotated-orbit}, this can prevent the planet being detected at high and low scattering angles. 
%
In this section, we derive the minimum ($\phi_\mathrm{min}$) and maximum ($\phi_\mathrm{max}$) scattering phase angles accessible for a given orbit of semi-major $a$ (in \si{\au}), line of sight inclination $i$, star distance $d_*$ in parsecs (\si{\parsec}) and for a given inner working angle of the instrument \IWA\ (in milliarcseconds). 
%
We define the inclination $i$, such that $i=0$ corresponds to a face-on orbit.

\begin{figure}
    \centering
    \includestandalone{tikz/annotated-orbit}
    \caption{
        Annotated orbit --- let's see if this works.
    }
    \label{fig:annotated-orbit}
\end{figure}

\begin{figure}%[t]
   \centering
   \includegraphics[width=0.99\columnwidth]{figures/orb-grid.pdf}
   \caption{
        The scattering phase angles accessible to a direct imaging mission can be limited by orbital inclination or by the coronagraphic inner working angle. 
        If the orbital inclination is close to face-on (top panels) then the scattering phase angles probed throughout an orbit are limited by inclination; for circular orbits accessible scattering phases are $\phi \in 90^\circ \pm i$. 
        For orbital inclinations close to edge-on (bottom panels) the planet goes through all phases, but the scattering phase angles near 0$^\circ$ and 180$^\circ$ are not accessible due to obscuration by the coronagraph mask; for circular orbits, the accessible scattering phases are $\phi \in 90^\circ \pm \arcsin({\rm IWA}\cdot d_*/a)$.
    }
    \label{fig:orb-grid}
    \script{plot-orb-grid.py}
\end{figure}

In the non-eccentric case (i.e., with a circular orbit and where the coronagraphic mask centred on the star), $\phi_\mathrm{min}$ and $\phi_\mathrm{max}$ are symmetric about 90 degrees (). 
We define the angle $\Delta \phi$ such that: 
\begin{equation}
    \label{eq:Delta_phi}
    \Delta \phi 
    = \phi_\mathrm{max} - \phi_\mathrm{min}
    =  2(\phi_\mathrm{max} - 90) 
    =  2(\phi_\mathrm{min} + 90)
\end{equation}

For a given system with inclination $i$, we have identified 2 regimes, shown in \cref{fig:orb-grid}:

@@

(2)~When parts of the disk (minimum and maximum scattering phase angles) are hidden by the focal plane mask, the scattering phase angle coverage does not depend on the inclination and only on the \IWA\ and orbit semi-major axis. 
In this section, we derive the analytical formula for this case in the circular orbit case.

%\begin{figure}%[t]
%    \centering
    % \includegraphics[width=\linewidth]{figures/scattering_angles.pdf}
%    \includestandalone{figures/scattering-phase-angle}
%    \caption{
%        The angular separation, $\varepsilon$, of a planet orbiting with semi-major-axis $a$ about a star a distance $d_*$ from the observer. 
%        The scattering phase angle is $\varphi$. 
%        The smallest accessible scattering phase angle occurs when the angular separation equals the coronagraph inner working angle: $\varphi_{\rm min} = {\rm IWA} \cdot d_*/a$.
%    }
%    \label{fig:scattering_angles}
%\end{figure}

%From \cref{fig:scattering_angles}, we deduce that for all points on the orbit, the scattering angle $\varphi$ satisfies:
\begin{equation}
    \sin(\varphi) = \frac{r_\mathrm{proj}}{a}
\end{equation}
with $r_\mathrm{proj}$ the on-sky projected planet distance (in \si{\au}), $a$ the distance to the star (constant in this section, in \si{\au}).

$\varepsilon$ is the angular separation of the planet at a given point of the orbit ($r_\mathrm{proj} = \delta d_*$). 
We access the minimum scattering angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
Using \cref{eq:Delta_phi}, we deduce:

\begin{equation}
    \cos\left(\dfrac{\Delta \phi}{2}\right) = \frac{\mathrm{IWA \cdot d_*}}{a}
\end{equation}

We now deduce the 2 regime equation for all cases (circular orbit): 
\begin{equation}
\label{eq:Delta_phi_max}
    \Delta \phi = 
    \begin{cases}
        2 \cdot i & \textrm{for} \cos(i) > \dfrac{\mathrm{IWA}\cdot d_* }{a}
  \\ 
        2 \cdot \cos^{-1}\left(\dfrac{\mathrm{IWA}\cdot d_* }{a}\right)  & \textrm{for} \cos(i) < \dfrac{\mathrm{IWA}\cdot d_* }{a}
    \end{cases}\,.
\end{equation}

\subsection{Edge-on circular orbits}
\label{sec:edge-on}
We initially performed a simple test considering only edge-on circular orbits. 
This gives an upper limit for circular orbits for the visible scattering phase range as a function of inclination. 
\Cref{fig:scatterplot} shows $\Delta \phi$ for the systems in the target list assuming a $3 \lambda / D$ coronagraph at $\SI{600}{\nano\meter}$. 
This optimistic case indicates that it will be possible to observe Rayleigh scattering for all targets, and that rainbows will be accessible in some cases.

\subsection{Assumed planet parameters}

As shown in \cref{sec:Delta_phi}, $\Delta \phi$ depends on the angular separation, inclination, and eccentricity of the planet's orbit. 
We use the angular separation for an Earth-Like planet with the same instellation as Earth, which is provided in the input star list. 
The inclination $i$ is sampled such that $\cos(i)$ follows a uniform distribution, $\cos(i) \sim \mathcal{U}(0, 1)$.

\subsection{Making Contrast Curves}
\textbf{Kim and David}
%
A second parameter that is crucial for detecting features like rainbows and ocean glint on Earth-like planets is the contrast, i.e. the relative flux of the reflected light from the planet divided by the stellar flux. 
While this is not the main focus of this paper, it is important to consider that the contrast and polarised contrast are not constant as function of scattering phase angles.
This is due to a combination of the changing illumination fraction of the exoplanets Earth facing surface and scattering effects. 
Here, we aim to present some simplified examples of the impact of the scattering angle dependent contrast on the detectability of rainbow and glint features. 
We note that these calculations are by no means complete or exact, given the many variables that go into modelling reflected fluxes and polarization fractions of Earth-like planet \citep{ treesstam2019,trees2022}.\\
The NASA Exoplanet Exploration Program’s Mission Star
List for the Habitable Worlds Observatory includes an estimated contrast ratio for an Earth twin in the habitable zone for every star. 
The contrast is calculated as 
\begin{equation}
C = F_p/F_* = p \phi (\alpha) (R_p/a)^2,
\end{equation}
where p is the geometric albedo, $\phi (\alpha)$ is the integral phase function at phase angle $(\alpha)$, $R_p$
is the planet’s radius, and $a$ is the separation of the planet from its star. 
For their calculations they made the simplified assumptions of circular orbits, a geometric albedo of $p=0.2$, and that the scattering phase function, $\phi (\alpha)$, can be described by a simple Lambertian reflectance phase function, e.g. isotropic scattering.
These calculations give a good order-of-magnitude calculation for the contrast of Earth-like planets around nearby stars.
Therefore, in this work we will assume that the calculated contrasts given in the list correspond to the reflected flux at quadrature and normalize the scattering phase functions to this point.
We note that we do not aim to accurately calculate the absolute scaling of the contrast as this depends on many factors and this would require to redo some of the work done for the star list.\\

We use the scattering phase functions from \cite{treesstam2019} for an Earth-like planet with an ocean surface with patchy clouds and a wind-speed of 7m/s at 670 nm.
The 670 nm is close to the center of the Rc band for Vega (642 nm). 
We normalize the reflected light curve at the value of 90 degrees (quadrature) and multiply it with the contrast.
In addition, we map the orbital phase for circular orbits to the scattering phase angle and the on-sky separation given the inclinations of 60 and 90 degrees.
Together they give the reflected light contrast as function of separation for a full orbit.
We multiply this flux with the degree of polarisation and retrieve the polarised contrast.

See Figure~\ref{fig:contrasts}.

\begin{figure*}%[t]
   \centering
   \includegraphics[width=0.99\textwidth]{figures/Contrast_vs_separation_inclination_angle_63.pdf}
   \caption{
    Contrast plots for three stars.
    }
    \label{fig:contrasts}
    \script{plot_contrasts.py}
\end{figure*}

@@

\subsection{Circular randomly inclined orbits}
\label{sec:circular}
For each star on the \hwo\ target list, we generate XX hypothetical habitable planets at the Earth-equivalent-flux semi-major axis. The orbital inclinations are randomly drawn from a distribution uniform in $\cos i$. 
We then use \cref{eq:Delta_phi_max} to determine the range of accessible scattering phase angles for each hypothetical planet. 
\todo{Figure X} shows the cumulative distribution for the most extreme scattering phase angles accessible for the \hwo target list, depending on the choice of inner working angle.

\textbf{Is there a plot for this?}

\subsection{Eccentric randomly inclined orbits}
\label{sec:eccentric}
We repeat the Monte Carlo simulation from \cref{sec:circular} assuming the orbital eccentricity follows a beta distribution with shape parameters $a=0.867$ and $b=3.03$ (cf. \citet{Guimond_2019} and references therein). 
We randomly drew 1000 eccentricities from the beta distribution for each star and determined the orbits. 
We then filtered on \IWA for multiples of 1, 2, 3 and 4 $\lambda / D$, before identifying which targets would be observable at the scattering phase angles where ocean glint, rainbows, or the Rayleigh peak occur. 
\Cref{fig:ball-o-yarn} shows a sample of the generated orbits with the filtered \IWA's. 
Each plot has been scaled such that the habitable zone occupies the same radius, hence the \IWA's vary in apparent size compared to the orbit.

@@

\begin{figure}
    \centering
    \includestandalone{tikz/scattering-angle}
    \caption{
        Scattering angle --- let's see if this works.
    }
    \label{fig:scattering-angle}
\end{figure}

@@

\subsection{Contrast Curves}
\label{sec:contrast}

\section{Discussion}
What did we determine? Can we actually detect this stuff for anything in this sample? Do we need to not something about things like A stars which are not included in the sample - they were rejected by the list creators - do we need to comment on this?  - we decided nope.

>>>>

\newcommand{\todo}[1]{\textcolor{red}{[#1]}}
\newcommand{\timmy}[1]{\textcolor{red}{[\textbf{Timmy:} #1]}}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory, a coronagraphic mission which would detect rocky planets in the habitable zone, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. The glint signature at scattering phases of $\sim$50--70$^\circ$ would be accessible to the Habitable Worlds Observatory in $\sim$20 systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in $\sim$55 systems, assuming a 60 mas inner working angle (\IWA). The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}

@@

%\textcolor{red}{introduce motivation, specs for habitable worlds observatory and the document we got the target list from.}

Surface liquid water is the definition of planetary habitability: water is required for life as we know it, and life will only be remotely detectable on an exoplanet if it is present at the surface. Hence the search for life on exoplanets typically focuses on habitable planets. To first order, the presence of surface liquid water can be predicted by its instellation---the flux it receives from its star. The habitable zone (HZ) describes the range of semi-major axes for which an Earth twin could maintain surface liquid water around a given main sequence star \cite{kasting93}.  The HZ is popular among planet hunters because a planet can be deemed habitable based on solely on its orbital separation, but it remains an untested hypothesis. Ideally, we will empirically establish the presence of liquid water on an exoplanet before trying to interpret its putative biosignatures.

\textcolor{blue}{Kim will enhance the polarimetry stuff, and we should break into multiple paragraphs.}\\
%\textbf{Why glint [Kim]}
% probe of surface conditions; habitability
\cite{2001Natur.412..885F} first noted the potential for time-resolved photometry and polarimetry to identify water oceans on the surface of an exoplanet. \textcolor{blue}{A sentence about rotational mapping here.} \cite{2006astro.ph.10518M} further quantified how photometric and polarimetric scattered light phase variations might betray the presence of oceans on terrestrial exoplanets.   
%--> No it was talked about previous to this, per the paper cited here in fact... "Stam & Hovenier(2006) independently examined the observability of the polarized signatures of an Earth-likeextrasolar planet, including Rayleigh scattering and sea-surface glint. Schneider (2006) had similar ideas." -KB
\cite{2008Icar..195..927W} simulated photometric and polarimetric phase curves of directly imaged terrestrial planets. They noted that surface oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases. \cite{Zugger_2010} simulated photometric and polarized phase variations for planets including the effects of partial cloud coverage and ocean waves. Although the presence of clouds and atmospheric Rayleigh scattering weaken the signal of an underlying ocean, they again find that oceans are detectable and distinguishable from other sources of polarization. \cite{2010ApJ...721L..67R} compared observations of the Earth's scattering phase curve to comprehensive radiative transfer simulations; they found that the photometric brightening of Earth at crescent phases is mostly due to forward scattering from clouds, with a relatively small contribution from ocean glint.
\citet{Zugger_2011} note that ocean glint might be observed in near infrared opacity windows, because of the reduced Rayleigh scattering at long wavelengths. This has the added advantage of minimizing the latitude--albedo effect, which could otherwise be a false-positive for photometric brightening at crescent phases due to glint \citep{2012ApJ...752L...3C}. \cite{2019A&A...626A.129T} simulated the photometric and polarimetric signatures of terrestrial planets with inhomogeneous clouds. \cite{2022A&A...664A.172T}...

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Observing scattering phenomena}

@@

\subsection{\hwo\ and Stellar Sample}
%\textcolor{blue}{Merge the 2 sections here into "Assumed technical specifications and sample"? I don't think this needs to be 2 subsections}
\hwo\ is still in the early stages of development, and the exact telescope and instrument design is undecided. 
%
In this work, we assume a \SI{6}{\meter} primary mirror, which is the currently favoured design for the observatory.
%
In addition, the observatory will utilise a coronagraph to suppress the stellar light and facilitate high contrast observations of their faint exoplanet companions. 
%
The \IWA\ of the coronagraph is currently undecided and will be driven by both the science requirements and technological limitations. 
%In this work, we investigate the scattering phase coverage obtainable for several choices for the \IWA\ of the observatory. 
%We also consider the Outer Working Angle (OWA) but for most targets this is not a limiting factor. 
For ease of comparison, \cref{tab:IWA_OWA} indicates the angular separations corresponding to multiples of $\lambda / D$ for a \SI{6}{\meter} telescope at a wavelength of $\lambda = \SI{-1}{\nano\meter}$ \todo{Update value!}.

@@

\hwo\ is currently envisioned to observe the stars stated in the NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory.%
\footnote{\url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list comprises $\sim$160 stars, the majority of which are Sun-like dwarfs; 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude of $R_c = 31$ and a contrast floor of \num{2.5e-11}. 
%with an occurrence rate of rocky planets in the optimistic habitable zone to be $\eta_{\oplus}$ = 0.24 \citep{decadal}. 
The corresponding adopted habitable zone limits are a semi-major axis of \SIrange{0.95}{1.67}{\au} for a solar twin, planet sizes between \SIrange{0.8}{1.4}{} Earth radii, and for non-solar stars, scale as square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}.

@@

The coronagraph can obscure part of the planet's orbit which, as demonstrated in \cref{fig:annotated-orbit}, can prevent the planet from being observed at high and low scattering angles. 
%
The range of scattering phase angles that can be observed will depend on the \IWA\ of the coronagraph as well as the shape and orientation of the exoplanet's orbit.

\begin{figure}
    \centering
    \includestandalone{tikz/annotated-orbit}
    \caption{
        The coronagraph can obscure the exoplanet at high and low scattering phase angles. This can prevent observations of scattering phenomena which could be used to indicate the habitability and atmospheric composition of the exoplanet. \todo{Add why scattering is cool} 
    }
    \label{fig:annotated-orbit}
\end{figure}

\subsubsection{Circular edge-on orbits}

In the non-eccentric case (i.e., with a circular orbit and where the coronagraphic mask centred on the star), minimum ($\phi_\mathrm{min}$) and maximum ($\phi_\mathrm{max}$) scattering phase angles accessible are symmetric about 90 degrees (). 
%
We define the angle $\Delta \phi$ such that: 
\begin{equation}
    \label{eq:Delta_phi}
    \Delta \phi 
    = \phi_\mathrm{max} - \phi_\mathrm{min}
    =  2(\phi_\mathrm{max} - 90) 
    =  2(\phi_\mathrm{min} + 90)
\end{equation}

The on-sky projected planet distance, $r_\mathrm{proj}$ (in \si{\au}) is shown in \cref{fig:scattering-angle} assuming a non-eccentric circular orbit with a semi-major $a$ (in \si{\au}) at a distance $d_*$ (in \si{\parsec}) from the observer. 
%
The scattering phase angle $\phi$ is this set up follows the equation,
\begin{equation}
    \sin(\phi) = \frac{r_\mathrm{proj}}{a}
\end{equation}
%
Using small angles, $r_\mathrm{proj} = \delta d_*$ where $\delta$ is the angular separation on-sky.
%
We access the minimum scattering angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
%
Using \cref{eq:Delta_phi}, we deduce:

\begin{equation}
    \label{eq:scattering_angle}
    \cos\left(\dfrac{\Delta \phi}{2}\right) = \frac{\mathrm{IWA \cdot d_*}}{a}
\end{equation}

\begin{figure}
    \centering
    \includestandalone{tikz/scattering-angle}
    \caption{
        The on-sky projected planet distance, $r_\mathrm{proj}$, of a planet with a semi-major $a$ at a distance $d_*$. The scattering phase angle, $\phi$, is the angle between the line of sight to the star and the line connecting the star and planet. $\delta$ is the on-sky angular separation as viewed by the observer. \todo{should the red angle be $\phi$?}
    }
    \label{fig:scattering-angle}
\end{figure}

\Cref{fig:scatterplot} shows $\Delta \phi$ for the systems in the target list assuming a $3 \lambda / D$ coronagraph at $\SI{600}{\nano\meter}$. 
%
This assumes each planet has a semi-major axis such that it receives an Earth-equivalent instellation, which is provided in the input star list.
%
For edge-on circular orbits, it will be possible to observe Rayleigh scattering for all targets and rainbows will be accessible in some cases.

\subsubsection{Circular inclined orbits}

Most systems will not be viewed as edge on from Earth.
%
For a given system with inclination $i$, we have identified 2 regimes, which are shown in \cref{fig:orb-grid}:

@@

(2)~When parts of the disk (minimum and maximum scattering phase angles) are hidden by the focal plane mask, the scattering phase angle coverage does not depend on the inclination and only on the \IWA\ and orbit semi-major axis.

\begin{figure}
   \centering
   \includegraphics[width=0.99\columnwidth]{figures/orb-grid.pdf}
   \caption{
        The scattering phase angles accessible to a direct imaging mission can be limited by orbital inclination or by the coronagraphic inner working angle. 
        If the orbital inclination is close to face-on (top panels) then the scattering phase angles probed throughout an orbit are limited by inclination; for circular orbits accessible scattering phases are $\phi \in 90^\circ \pm i$. 
        For orbital inclinations close to edge-on (bottom panels) the planet goes through all phases, but the scattering phase angles near 0$^\circ$ and 180$^\circ$ are not accessible due to obscuration by the coronagraph mask; for circular orbits, the accessible scattering phases are $\phi \in 90^\circ \pm \arcsin({\rm IWA}\cdot d_*/a)$.
    }
    \label{fig:orb-grid}
    \script{plot-orb-grid.py}
\end{figure}

Using \cref{eq:scattering_angle}, we deduce the 2 regime equation for inclined circular orbits: 
\begin{equation}
\label{eq:Delta_phi_max}
    \Delta \phi = 
    \begin{cases}
        2 \cdot i & \textrm{for} \cos(i) > \dfrac{\mathrm{IWA}\cdot d_* }{a}
  \\ 
        2 \cdot \cos^{-1}\left(\dfrac{\mathrm{IWA}\cdot d_* }{a}\right)  & \textrm{for} \cos(i) < \dfrac{\mathrm{IWA}\cdot d_* }{a}
    \end{cases}\,.
\end{equation}

\subsubsection{Eccentric orbits}

\todo{how was the angle coverage calculated in the monte carlo?}

\timmy{I think so far, the MC code doesn't compute angle coverage. I came up with this quick and dirty sketch to help us get started:}

\includegraphics[width=\linewidth]{figures/ellipse.pdf}

\timmy{If you agree that this makes sense, I could try to add this to the yarn ball script by Matt?}

%The inclination $i$ is sampled such that $\cos(i)$ follows a uniform distribution, $\cos(i) \sim \mathcal{U}(0, 1)$.

\subsection{Simulating contrast curves}
\textbf{Kim and David}
%
A second parameter that is crucial for detecting features like rainbows and ocean glint on Earth-like planets is the contrast, i.e. the relative flux of the reflected light from the planet divided by the stellar flux. 
%
While this is not the main focus of this paper, it is important to consider that the contrast and polarised contrast are not constant as function of scattering phase angles.
%
This is due to a combination of the changing illumination fraction of the exoplanets Earth facing surface and scattering effects.
%
Here, we aim to present some simplified examples of the impact of the scattering angle dependent contrast on the detectability of rainbow and glint features. 
%
We note that these calculations are by no means complete or exact, given the many variables that go into modelling reflected fluxes and polarization fractions of Earth-like planet \citep{ treesstam2019,trees2022}.

The NASA Exoplanet Exploration Program’s Mission Star List for the Habitable Worlds Observatory includes an estimated contrast ratio for an Earth twin in the habitable zone for every star. 
%
The contrast is calculated as 
\begin{equation}
C = F_p/F_* = p \phi (\alpha) (R_p/a)^2,
\end{equation}
where p is the geometric albedo, $\phi (\alpha)$ is the integral phase function at phase angle $(\alpha)$, $R_p$ is the planet’s radius, and $a$ is the separation of the planet from its star. 
%
For their calculations they made the simplified assumptions of circular orbits, a geometric albedo of $p=0.2$, and that the scattering phase function, $\phi (\alpha)$, can be described by a simple Lambertian reflectance phase function, e.g. isotropic scattering.
%
These calculations give a good order-of-magnitude calculation for the contrast of Earth-like planets around nearby stars.
%
Therefore, in this work we will assume that the calculated contrasts given in the list correspond to the reflected flux at quadrature and normalize the scattering phase functions to this point.
%
We note that we do not aim to accurately calculate the absolute scaling of the contrast as this depends on many factors and would have required some of the work for the star list to be redone.

We use the scattering phase functions from \cite{treesstam2019} for an Earth-like planet with an ocean surface with patchy clouds and a wind-speed of $\SI{7}{\meter\per\second}$ at $\SI{670}{\nano\meter}$.
%
The $\SI{670}{\nano\meter}$ is close to the center of the Rc band for Vega ($\SI{642}{\nano\meter}$). 
%
We normalize the reflected light curve at the value of 90 degrees (quadrature) and multiply it with the contrast.
%
In addition, we map the orbital phase for circular orbits to the scattering phase angle and the on-sky separation given the inclinations of $60$ and $90$ degrees. \todo{Only the 60 degree plot is here at the moment}
%
Together they give the reflected light contrast as function of separation for a full orbit.
%
We multiply this flux with the degree of polarisation and retrieve the polarised contrast.
%
The resulting contrast curves are shown in \cref{fig:contrasts} for three of the closest target stars, the G dwarf $\alpha$ Cen A, the K dwarf $\epsilon$ Eri and the M dwarf Lalande 21185.

\begin{figure*}%[t]
   \centering
   \includegraphics[width=0.99\textwidth]{figures/Contrast_vs_separation_inclination_angle_63.pdf}
   \caption{
    The reflected light contrast and orbital separation of an Earth-like planet with an ocean surface and patchy clouds over a planetary orbit assuming an orbital inclination of $60^\circ$ for the stars $\alpha$ Cen A, $\epsilon$ Eri and Lalande 21185. The solid line indicates the contrast in unpolarised light with the contrast at quadrature marked by a solid dot. The polarised component is indicated by the colored dots for which the color represents the scattering phase angle from quadrature. The light grey points show the quadrature contrasts of the other targets in the star list and the dashed lines indicate $1,2$ and $3$ times the \IWA\ for \hwo.
    }
    \label{fig:contrasts}
    \script{plot_contrasts.py}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Circular randomly inclined orbits}
\label{sec:circular}

For each star on the \hwo\ target list, we generate XX hypothetical habitable planets at the Earth-equivalent-flux semi-major axis. 
%
The orbital inclinations are randomly drawn from a distribution uniform in $\cos i$. 
%
We then use \cref{eq:Delta_phi_max} to determine the range of accessible scattering phase angles for each hypothetical planet.
%
\todo{Figure X, where the plot for this?} shows the cumulative distribution for the most extreme scattering phase angles accessible for the \hwo\ target list, depending on the choice of inner working angle.

\todo{what does this plot show, how many systems could we see?}

\subsection{Eccentric randomly inclined orbits}
\label{sec:eccentric}
We repeat the Monte Carlo simulation from \cref{sec:circular} assuming the orbital eccentricity follows a beta distribution with shape parameters $a=0.867$ and $b=3.03$ (cf. \citet{Guimond_2019} and references therein). 
%
We randomly drew 1000 eccentricities from the beta distribution for each star and determined the orbits. 
%
We then filtered on \IWA\ for multiples of 1, 2, 3 and 4 $\lambda / D$, before identifying which targets would be observable at the scattering phase angles where ocean glint, rainbows, or the Rayleigh peak occur. 
%
\Cref{fig:ball-o-yarn} shows a sample of the generated orbits with the filtered \IWA's. 
%
%Each plot has been scaled such that the habitable zone occupies the same radius, hence the \IWA's vary in apparent size compared to the orbit.

\todo{Is there a plot for this? How many systems could we see? Why is it different to non-eccentric cases?}

\timmy{No plot / results yet; see above.}

@@

@@

%\subsection{Contrast Curves}
%\label{sec:contrast}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusions}
%What did we determine? 
%Can we actually detect this stuff for anything in this sample? 
%Do we need to not something about things like A stars which are not included in the sample - they were rejected by the list creators - do we need to comment on this?  - we decided nope.

The Habitable Worlds Observatory is a coronagraphic space mission envisioned to detect and characterise Earth-like exoplanets in the 2040s \todo{check}. 
%
The instrument design and capabilities are still under development and thus can be driven by the science goals of the mission.
%
In this work, we investigated the possibility of using \hwo\ to detect different scattering phenomena which can be used to indicate the habitability of an exoplanet. \todo{more of why this is cool}
